|

### 2022-06-10 paper.Rmd
bring Rmd in line with accepted version

---
journal: International Journal of Human-Computer Studies
title: "The Goldilocks Level of Friction: Using User Numbers, Ratings, and Reviews to Evaluate Digital Self-Control Tools"
keywords: Attention, self-control, self-regulation, distraction, digital wellbeing, focus
abstract: |
  Much effort has been invested in designing digital systems that keep people 'hooked'. By contrast, comparatively little is known about how designers can support people in re-gaining control. Online, however, hundreds of apps and browser extensions promise to help people self-regulate use of digital devices. Reviews and popularity metrics for these digital self-control tools (DSCTs) can indicate which design patterns are useful in the wild. Moreover, they reveal how platforms such as Android and iOS differ in the ecosystems they enable for DSCTs, which has important practical implications for end users. We analysed reviews, installation numbers, and ratings for 334 DSCTs on the Google Play, Chrome Web, and Apple App stores, investigating what user reviews reveal about usage contexts and key design challenges, and how functionality relates to popularity metrics. Our thematic analysis of 1,529 reviews (sampled from a data set of 54,320 reviews scraped in March 2019) found that DSCTs are seen as highly important to stay focused on less instantly rewarding tasks when digital distractions are easily available. Users seek DSCTs that adapt to their personal definitions of 'distracting use', and which provide incentives sufficient to change behaviour without feeling overly intrusive. DSCTs' interventions can elicit negative emotions, which appeared to be an important reason that the DSCT accurately capture users' definition of distraction. Tools that combine two or more types of design patterns (e.g., website blocking and goal reminders) received slightly higher ratings than those implementing a single type. This mirrored suggestions in the reviews to combine different types of design patterns to provide a incentive for behaviour change that is 'just right'. 
  UPDATE FINAL SENTENCE AFTER REWRITE: We discuss implications for research and design, including how basic psychological research on self-regulation can inform evaluation measures for DSCTs, as well as guide future exploration of how design patterns and framings of 'distraction' interact.
bibliography: references.bib

@@

The constant connectivity of smartphones and laptops provides innumerable benefits, but also unprecedented opportunities for distraction.
Indeed, as the business models of many tech companies have come to rely on nudging people into using digital services extensively (cf. the 'attention economy' [@davenportAttentionEconomyUnderstanding2001; @einsteinBlackOpsAdvertising2016; @wuAttentionMerchantsEpic2016]), many people routinely struggle to exercise self-control over their device use [@coxDesignFrictionsMindful2016; @koNUGUGroupbasedIntervention2015; @marottaOnlineDistractionsWebsite2017; @tranModelingEngagementDisengagementCycle2019; @hinikerMyTimeDesigningEvaluating2016].
As a result, public debate about negative effects of 'persuasive technologies' has grown [@OptimizingEngagementUnderstanding2019; @orlowski2020], alongside increased political pressure on technology platforms to safeguard users [@allpartyparliamentarygrouponsocialmediaandyoungpeoplesmentalhealthandwellbeingNewFiltersManageImpact2019].

On online stores, a flourishing market exists for 'digital self-control tools' (DSCTs), where hundreds of apps and browser extensions promise to help people take back control over their device use.
These tools offer interventions such as blocking apps or removing distracting website elements, tracking and visualising usage, or providing rewards or punishments for how devices are used [@lyngsSelfControlCyberspaceApplying2019].
Similarly, Apple and Google now offer various usage logging and self-limiting features in their 'ScreenTime' and 'Digital Wellbeing' tools, respectively [@appleIOS12Introduces2018; @gonzalezResearchGoogleNew2018; @googleDigitalWellbeing2018], and even Facebook have introduced tools 'for managing your time on Facebook and Instagram' [@ranadiveNewToolsManage2018].

In academia, a growing amount of HCI research is similarly focused on developing and evaluating various design patterns that help users self-regulate digital device use [@lottridgeBrowserDesignImpacts2012; @kimGoalKeeperExploringInteraction2019; @pinderPushAwaySmartphone2019; @tsengOvercomingDistractionsTransitions2019; @mongeroffarelloRaceDigitalWellbeing2019; @lyngsSelfControlCyberspaceApplying2019][^1].
This work has gradually built an understanding of how a range of design patterns from visualisations of time spent [@whittakerDonWasteMy2016] to goal-setting with social support [@koNUGUGroupbasedIntervention2015] can help people stay in control.
However, we are still in the early days of answering basic questions including how to define and measure 'success' for these design patterns, how implementation details of specific design patterns --- such as the level of 'friction' --- influence their usefulness, and how much people vary in how useful they find particular interventions [@kimGoalKeeperExploringInteraction2019; @markEffectsIndividualDifferences2018; @tseng2019; @lyngsTellMeWhat2018; @hinikerMovingGreatScreen2019].

[^1]: We use the term 'design patterns' to refer to specific functions intended to support people in self-regulating their digital device use (e.g., hiding newsfeeds, providing goal reminders).
    We use the term 'digital self-control tool' (DSCT) to refer to apps and browser extensions that implement a given design pattern (e.g., a browser extension for hiding Facebook's newsfeed).
    We occasionally use the term 'intervention' to refer to the implementation of a design pattern in the context of it changing a user's usual digital environment, or when it is being evaluated in a research study.

To answer such questions, most existing studies have taken an experimental approach and compared user behaviour and perceptions when applying a given intervention to its absence and/or some alternative intervention [@lyngs2021].
Yet, while controlled studies are appropriate and feasible for evaluating and comparing a small number of design patterns, they are difficult to scale to broadly assess a large numbers of patterns and implementations.

A complementary approach, which can help scope the range of design patterns and implementations to explore in controlled studies, is to investigate the existing landscape of DSCTs on app and web stores [@lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018].
Here, user reviews and popularity metrics can provide information about user needs, contexts of use, and the relative utility of different patterns and implementations 'in the wild'.
However, few studies to date have taken this approach.
Hence, in the present paper, we contribute to the HCI literature on design patterns for digital self-control by using app and web store information as a resource for understanding user self-experimentation with DSCTs.
Specifically, we present an analysis of user reviews, functionality, installation numbers, and ratings for 334 digital self-control tools, drawn from Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] who reviewed functionality in such tools on the Google Play, Chrome Web, and Apple App stores.

Our analysis of user reviews finds that digital self-control tools are seen as highly important to supports focus on less instantly rewarding goals when digital distractions are easily available, specifically in contexts related to productivity (including work, studying, reading, and writing), and going to sleep.
DSCTs appeared to be especially useful for people who describe themselves as 'addicts' or as struggling with distraction due to conditions like ADHD.

Users want DSCTs to adapt to their personal definition of 'distraction', as seen in common requests for customisation of *what* and *when* particular usage patterns should be blocked or tracked.
One important reason that users' and tools' conceptualisation of distraction should be in close alignment is that DSCTs' interventions, useful as they may be, can elicit negative emotional reactions.
This may cause the effects of interventions to be counterproductive, especially if users perceive DSCTs' goals or framing to differ from their own.

Finally, users want DSCTs to provide carefully calibrated incentives that are sufficient to incentivise intended behaviour change without feeling overly intrusive.
In our analysis, DSCTs 'Goldilocks level'[^2] varied between reviewers.
However, reviewers commonly wanted tools that provided incentives strong enough to keep themselves on track in moments of temptation or absentmindedness.
Overly strong incentives could, however, backfire if they made users feel like a failure (in the case of reward/punishment tools) or did not allow some way to handle genuine emergency use (in the case of block/removal tools).

[^2]: After the fairy tale 'Goldilocks and the Three Bears', where a young girl tastes three different bowls of porridge and prefers the one that is neither too hot nor too cold, but 'just right'.
    The concept of the 'just right amount' is often referred to as the 'Goldilocks principle'.
    For example, a widely cited paper on the relationship between screen time and mental well-being documented that a balance point in between "too much" and "too little" tech use seemed optimal, which they referred to as the 'digital Goldilocks hypothesis' [@Przybylski2017].

One way for tools to provide properly calibrated incentives might be by carefully combining different types of design patterns (e.g., using both block/removal and goal-advancement features).
This was often asked for in reviews' feature suggestions, and tentatively supported by higher overall ratings given to tools that combine two or more types compared to just one.
A final consideration related to the amount of effort involved in operating a tool, where low friction to set up and initiate use (e.g., via schedules where a tool automatically turns on), but high friction to override it, was important.

@@

In HCI, a range of studies have developed and evaluated design patterns for digital self-control, ranging from ambient visualisations of time spent [@whittakerDonWasteMy2016] and distraction blocking [@kimGoalKeeperExploringInteraction2019] to goal-setting and social support [@koNUGUGroupbasedIntervention2015].
This research is starting to build an evidence base of how specific interventions to support self-regulation influence user behaviour and perceptions, but basic questions remain open.
These include *conceptual challenges* (e.g., what metrics digital self-control tools should seek to optimise, as simply measuring impact on 'screen time' is deemed inadequate by most researchers [@cecchinatoDesigningDigitalWellbeing2019; @hinikerMovingGreatScreen2019; @lukoffDigitalWellbeingWay2019]); related gaps in our understanding of *individual variation in user needs* (such as how how widely people differ in what they consider 'distracting' usage [@cecchinatoDesigningDigitalWellbeing2019; @lukoffDigitalWellbeingWay2019]); and *practical implementation challenges* (such as how much 'friction' tools should apply to hold users accountable for their past preferences [@kimGoalKeeperExploringInteraction2019; @coxDesignFrictionsMindful2016; @lyngsTellMeWhat2018]).

One approach to addressing these questions is to evaluate new or existing interventions in lab and field studies, and/or conduct surveys or interviews with users.
A complementary approach, which we explore in the present paper, is to investigate the landscape of digital self-control tools available online.
Here, many hundreds of tools are exploring a wide range of strategies, from limiting the amount of functionality available on home screens (e.g.\~the Android app \emph{LessPhone Launcher} [@mohanLessPhoneLauncher2019]) and hiding or modifying content on distracting websites (e.g. *Newsfeed Eradicator* [@jdevNewsFeedEradicator2019]), to gamifying self-control by tying device use to the wellbeing of virtual creatures (e.g. *Forest* [@seekrtechForestStayFocused2018]).

Some of these tools have millions of users [@lyngsSelfControlCyberspaceApplying2019], and they effectively represent a myriad of natural studies in which individuals self-experiment with tools that represent not only one or more design patterns, but specific ways to implement those patterns [@daskalovaSleepCoacher2018; @leeSelfExperimentation2017].
Indicators of the outcome of these experiments include user reviews, store ratings, and installation numbers, which may provide valuable information about the usefulness of different design patterns and implementations.

Other areas of HCI research have benefited from taking such an approach.
For example, in mental health research, thematic analysis of user reviews has been used to explore factors important for optimising user experience and support engagement in apps for cognitive behavioural therapy [@bakkerMentalHealthSmartphone2016; @huguetSystematicReviewCognitive2016; @luiEvidencebasedAppsReview2017; @sucalaAnxietyThereApp2017; @stawarzUserExperienceCognitive2018].
<!-- Similar research found that content in popular apps marketed as reducing symptoms of anxiety and worry was largely inconsistent with evidence-based treatments [@kertzReviewFreeIPhone2017]. --> <!-- This opened avenues for further research including how established CBT practices could more effectively be disseminated and, conversely, what might be learned from alternative approaches that proved popular in the wild. --> <!-- HCI research on design patterns for digital self-control might in a similar way derive actionable insights from the large number of tools for this purpose available online. --> However, few studies on digital self-control have so far taken this route: To the best of our knowledge, four published studies have described and/or categorised functionality in samples of digital self-control tools available online [@biskjaerTakingActionDistraction2016; @lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018]: Velthoven et al. [@vanvelthovenProblematicSmartphoneUse2018] presented aims and features in 21 tools identified via the software recommendation platform [alternativeto.net](https://alternativeto.net); Biskjær et al. [@biskjaerTakingActionDistraction2016] presented a taxonomy of functionality based on 10 tools from the Google Play, Chrome Web, and Apple App stores, and online tech magazines; Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019] presented a taxonomy of functionality based on 42 mobile apps from the Google Play store; and Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] presented a taxonomy of functionality based on 367 tools from the Google Play, Chrome Web and Apple App stores, which was mapped to psychological mechanisms in a dual systems model.

Only one of these studies analysed user reviews, namely Roffarello & Russis [@mongeroffarelloRaceDigitalWellbeing2019], which included thematic analysis of 1,128 reviews from 42 smartphone apps, and none incorporated popularity metrics.
Moreover, Roffarello & Russis' analysis was restricted to 42 smartphone apps drawn from the Google Play store and did not specify how the reviews were distributed across apps with different functionality.
Thematic analysis is likely to yield different results depending on how reviews are sampled across tools with different functionality, and depending on the analysis strategy [@chenARminerMiningInformative2014; @braunThematicAnalysis2018].
Therefore, additional analyses with broader sampling of tools, and transparent sampling of reviews from those tools, might be helpful.

On this background, we extended existing studies of publicly available digital self-control tools by (i) scraping installation numbers, ratings, and user reviews from a broad sample of tools across stores and types of functionality, (ii) analysing review content using 'reflexive' thematic analysis [@braunThematicAnalysis2018], and (iii) analysing popularity metrics' relationship to tools' design features, and (iv) making our materials, analysis pipelines and data openly available for the purpose of transparency and reusability [@harozUpdatesCHISubmission2019].

@@

Data, analysis scripts, and materials for the present paper are available at [osf.io/cdqub/](https://osf.io/cdqub/).

We investigated the 367 digital self-control tools reviewed by Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019], as this paper provided the largest currently available review of DSCTs.
Lyngs et al. identified these tools by downloading search results for the terms 'distraction', 'addiction', 'motivation', and 'self-control', individually as well as prefixed by 'internet', 'smartphone', and 'laptop', on the Chrome Web, Google Play, and Apple App stores.
From the initial results (4890 distinct apps and extensions), they included tools explicitly designed to help people self-regulate digital device use, resulting in a data set of 367 tools (222 browser extensions from Chrome Web, 87 apps from Google Play, and 58 apps from the Apple App store).[^3]

We adopted Lyngs et al.'s classification of functionality within these tools, which clustered their features into four types of design patterns for digital self-control:

-   \emph{block/removal} (features for blocking distractions, such as temporarily locking the user out of specific apps, or for removing them in the first place, such as hiding recommended videos on YouTube)
-   \emph{self-tracking} (features for tracking and/or visualising device use, such as showing how much time the user has spent in particular apps or on their devices overall, or letting the user start a countdown timer when they need to focus)
-   \emph{goal-advancement} (features for reminding the user of their usage goals, such as displaying pop-ups when a specific amount of time has been spent, or replacing the content of new browser tabs with to-do lists)
-   \emph{reward/punishment} (features that add incentives for using devices in particular ways, such as rewarding the user with points/streaks when they do not use their phone, or specific apps)

@@

In March 2019, we scraped additional information about these tools, including installation numbers, ratings, and store descriptions, as well as publicly available user reviews.
For apps on the Google Play store, we used a pre-existing script [@olanoGoogleplayscraper2018] to scrape tool information and reviews.
For apps on the Apple App store, we used a pre-existing script [@olanoAppstorescraper2018] to scrape reviews, and used our own script (using the R packages \href{https://rvest.tidyverse.org}{\texttt{rvest}} and \href{http://ropensci.github.io/RSelenium/}{\texttt{RSelenium}}) to scrape tool information.
As the pre-existing script sometimes failed to retrieve all reviews visible on tools' pages on the Apple App store, we wrote an additional R script to scrape reviews when manual inspection found some to be missing.
For browser extensions on the Chrome Web store, we wrote our own R script to scrape both tool information and user reviews.

@@

All stores displayed average ratings, as well as a count of how many ratings a tool had received.
For installation numbers, the stores provided different information: the Chrome Web store provided an exact number of users, and the Google Play store provided a 'minimum number of installs' (e.g. '100,000+').
The Apple App store provided no direct information about user numbers.
For processing which ranked tools by installation numbers, we used number of ratings as a proxy for relative numbers of users for tools on the Apple App store.

\subsection{Sampling and thematic analysis of user reviews}

To get a broad coverage of reviews for tools across stores and types of design patterns, we used an R script to systematically sample user reviews for qualitative analysis from the full scraped data.
To ensure that the reviews we analysed came from tools with a minimum level of engagement from users, we excluded reviews from tools on the Google Play and Chrome Web stores that had fewer than 100 installations/users, and tools from the Apple App Store that had fewer than 5 ratings.
To make the sampled reviews more likely to be informative, we also excluded reviews shorter than 10 words.
Our script then iterated through the four types of design patterns described above, on each iteration picking the tool on each store that included that specific type of design pattern and had the highest number of users (Google Play and Chrome Web) or ratings (Apple App store).
The script then randomly sampled up to 30 user reviews from each of those tools (if a specific tool had fewer than 30 reviews, all of its reviews would be included), and started over.
The script ran until more than 1,500 reviews were sampled in total.
A tool could only be sampled from once.

Next, we conducted inductive thematic analysis of the sampled reviews, following the 'reflexive' approach described in [@braunUsingThematicAnalysis2006] and [@braunThematicAnalysis2018].
Our work was informed by an earlier analysis of 961 reviews randomly sampled from the same data set by the lead author of the present paper [@lyngs2021a, ch. 4].
However, our analysis in the present paper was conducted from scratch, with a new sample of user reviews drawn in the manner described.
First, the lead author (UL) read through all user reviews sampled for thematic analysis and conducted initial coding of recurrent patterns relevant to the research topic.
A second author (LC) read through one third of the reviews and independently coded recurrent patterns of meaning.
Afterwards, these two authors read through the coded excerpts, recoded them, iteratively sorted the codes into potential themes, and discussed emerging themes.

Thematic coding was conducted using [NVivo](https://www.qsrinternational.com/nvivo-qualitative-data-analysis-software/home/) v1.6.1; all quantitative analyses were conducted in R.

@@

At the time of scraping, `r num_missing` of the tools from Lyngs et al.'s review had become unavailable.
Hence, we scraped information for `r nrow(meta_data)` tools: `r num_w_info_by_store %>% filter(store == "chrome") %>% pull(n)` extensions from Chrome Web, `r num_w_info_by_store %>% filter(store == "play") %>% pull(n)` apps from Google Play, and `r num_w_info_by_store %>% filter(store == "apple") %>% pull(n)` apps from the Apple App store.

@@

(ref:word-cloud-caption) Word cloud depicting frequencies of terms across all `r format(nrow(reviews_w_main_cat_data), big.mark = ",")` reviews. Font size indicates relative frequency. The word cloud excludes the terms 'app/s', 'phone', and 'extension', as well as stop words (i.e., common English words unlikely to be informative such as "the", "of", "to", etc. Stop words were excluded using the R package `tidytext` [@silge2016tidytext]). Colouring is aesthetic and does not map to any characteristics of the data.

@@

```{r, include=FALSE}
(num_reviews_summary <- reviews_w_main_cat_data %>% 
  group_by(tool_id) %>% 
  summarise(num_reviews = n()) %>% 
  summarise(median = median(num_reviews),
            mean = mean(num_reviews),
            min = min(num_reviews),
            max = max(num_reviews),
            IQR = IQR(num_reviews)))

(word_count_summary <- reviews_w_main_cat_data %>% 
  summarise(median = median(review_word_count),
            mean = mean(review_word_count),
            min = min(review_word_count),
            max = max(review_word_count),
            IQR = IQR(review_word_count)))

>>>>

---
journal: International Journal of Human-Computer Studies
title: |
  The Goldilocks Level of Support:\
  Using User Reviews, Ratings, and Installation Numbers to Investigate Digital Self-Control Tools Tools
keywords: Attention, distraction, digital self-control, digital wellbeing, focus, Goldilocks principles, self-regulation
abstract: |
  Much effort has been invested in designing digital systems that keep people 'hooked'. By contrast, comparatively little is known about how designers can support people in re-gaining control. Online, however, hundreds of apps and browser extensions promise to help people self-regulate use of digital devices. Reviews and popularity metrics for these digital self-control tools (DSCTs) can indicate which design patterns are useful in the wild. Moreover, they reveal how platforms like Android and iOS differ in the ecosystems they enable for DSCTs, which has important implications for end users. We analysed reviews, installation numbers, and ratings for 334 DSCTs on the Google Play, Chrome Web, and Apple App stores, investigating what user reviews reveal about usage contexts and key design challenges, and how functionality relates to popularity metrics. Our thematic analysis of 1,529 reviews (sampled from a data set of 53,978 distinct reviews scraped in March 2019) found that DSCTs are seen as highly important for focusing on less instantly rewarding tasks when digital distractions are easily available. Users seek DSCTs that adapt to their personal definitions of distraction, and provide support that is sufficient to change behaviour without feeling too coercive. Reviewers suggested combining design patterns to provide a level of support that is 'just right'. This was mirrored in the ratings, where tools combined different types of design patterns (e.g., website blocking and goal reminders) tended to receive higher ratings than those implementing a single type. We discuss implications for research and design, including how design patterns in DSCTs interact, and how psychological reactance to DSCTs can be reduced.

bibliography: references.bib

@@

The constant connectivity of smartphones and laptops provides innumerable benefits, but also unprecedented opportunities for distraction. 
Indeed, as the business models of many tech companies have come to rely on nudging people into using digital services extensively (cf. the 'attention economy' [@davenportAttentionEconomyUnderstanding2001; @einsteinBlackOpsAdvertising2016; @wuAttentionMerchantsEpic2016]), many people routinely struggle to exercise self-control over their device use [@coxDesignFrictionsMindful2016; @koNUGUGroupbasedIntervention2015; @marottaOnlineDistractionsWebsite2017; @tranModelingEngagementDisengagementCycle2019; @hinikerMyTimeDesigningEvaluating2016]. 
In HCI, a growing amount of research attempts to address this challenge by developing and evaluating design patterns that help users self-regulate[^1] [@lottridgeBrowserDesignImpacts2012; @kimGoalKeeperExploringInteraction2019; @pinderPushAwaySmartphone2019; @tsengOvercomingDistractionsTransitions2019; @mongeroffarelloRaceDigitalWellbeing2019; @lyngsSelfControlCyberspaceApplying2019].
This work is gradually building an understanding of how interventions ranging from ambient visualisations of time spent [@whittakerDonWasteMy2016] and distraction blocking [@kimGoalKeeperExploringInteraction2019] to goal-setting with social support [@koNUGUGroupbasedIntervention2015] can help people stay in control.

Most of this research has taken an experimental approach and compared user behaviour and perceptions when applying a given intervention to its absence and/or some alternate intervention [@lyngs2021]. 
In the present paper, we take an alternative approach. On online stores for apps and browser extensions, hundreds of 'digital self-control tools' (DSCTs) provide similar interventions that block apps or hide distracting website elements, track and visualise use, provide rewards or punishments for how devices are used, and much more [@lyngsSelfControlCyberspaceApplying2019]. User reviews and popularity metrics from these tools might provide valuable complementary information about the utility of different design patterns and implementations 'in the wild', and help scope future controlled studies. However, a limited number of studies have investigated what we might learn from this source of data.

Hence, in this paper, we explore how user reviews and other store information can contribute to HCI research on design patterns for digital self-control. Specifically, we analyse user reviews, functionality, installation numbers, and ratings for 334 DSCTs on the Google Play, Chrome Web, and Apple App stores (drawn from [@lyngsSelfControlCyberspaceApplying2019]), investigating user needs, contexts and purposes of use, and key design challenges.

Our analysis of user reviews found that DSCTs are seen as highly important to support focus on less instantly rewarding tasks when digital distractions are easily available, especially in contexts related to productivity (including work, studying, reading, and writing), and going to sleep. DSCTs seemed particularly useful for reviewers describing themselves as 'addicts' or otherwise struggling intensely with distraction, for example due to conditions like ADHD.

Reviewers wanted DSCTs to flexibly adapt to their personal definition of 'distracting use', as seen in common requests for customisation of *what* and *when* particular usage patterns should be blocked or tracked. One important reason was that DSCTs' interventions, useful as they may be, can elicit negative emotional reactions. This could cause their interventions to be counterproductive, especially if users perceived DSCTs' framing of distracting use to differ from their own.

Finally, reviewers wanted DSCTs to provide carefully calibrated support that is sufficient to support intended behaviour change without feeling too coercive. This 'Goldilocks level'[^2] varied between reviews, but most wanted DSCTs to provide support powerful enough to keep the user on track in moments of temptation or absentmindedness. Thus, blocking tools were often criticised if they were too easy to override because they could then fail to hold people accountable in moments of temptation. On the other hand, DSCTs were also criticised if their incentives were *too* strong, for example if blocking tools did not include some way to handle genuine emergency use, or if tools providing reward/punishment made users feel dejected when they failed to use their devices as intended.

One way DSCTs might provide properly calibrated support is by combining different types of design patterns (e.g., combining distraction blocking with goal-reminders). This was sometimes requested in the reviews, and tentatively supported by rating scores, where tools that combine two or more types of design patterns received higher ratings than those implementing a single type.

@@

In recent years, HCI researchers have started to build an understanding of how specific design patterns intended to support self-regulation over digital device can influence user behaviour and perceptions [@whittakerDonWasteMy2016; @kimGoalKeeperExploringInteraction2019; @kimLocknTypeLockoutTask2019; @kovacsRotatingOnlineBehavior2018]. However, basic questions still remain open. These include *conceptual challenges* (e.g., what outcome measures we should seek to optimise, as simply measuring impact on 'screen time' is deemed inadequate by most researchers [@cecchinatoDesigningDigitalWellbeing2019; @hinikerMovingGreatScreen2019; @lukoffDigitalWellbeingWay2019]); related gaps in our understanding of *individual variation in user needs* (such as how how widely people differ in what they consider 'distracting' use [@cecchinatoDesigningDigitalWellbeing2019; @lukoffDigitalWellbeingWay2019]); and *practical implementation challenges* (such as how much 'friction' tools should apply to hold users accountable for their past preferences [@kimGoalKeeperExploringInteraction2019; @coxDesignFrictionsMindful2016; @lyngsTellMeWhat2018; @kimLocknTypeLockoutTask2019]).

One approach to addressing these questions is to evaluate new or existing interventions in lab and field studies. While this is appropriate and feasible for evaluating and comparing a small number of design patterns, it is difficult to scale to broadly assess a large numbers of patterns and implementations.

A complementary approach is to investigate the landscape of DSCTs available online. Here, hundreds of tools are exploring a wide range of strategies, from hiding or modifying content on distracting websites (e.g. *Newsfeed Eradicator* [@jdevNewsFeedEradicator2019]) or limiting the amount of functionality available on home screens (e.g. the Android app *LessPhone Launcher* [@mohanLessPhoneLauncher2019]), to gamifying self-control by tying device use to the wellbeing of virtual creatures (e.g. *Forest* [@seekrtechForestStayFocused2018]). Some of these tools have millions of users [@lyngsSelfControlCyberspaceApplying2019], and they effectively amount to a myriad of natural studies in which individuals self-experiment with tools that represent not only one or more design patterns, but specific ways to implement those patterns [@daskalovaSleepCoacher2018; @leeSelfExperimentation2017]. Indicators of the outcome of these experiments include user reviews, store ratings, and installation numbers, which can provide valuable information about the usefulness of different design patterns and implementations.

Other areas of HCI research have benefited from such an approach. For example, in mental health research, thematic analysis of user reviews has been used to explore factors important for optimising user experience and support engagement in apps for cognitive behavioural therapy [@bakkerMentalHealthSmartphone2016; @huguetSystematicReviewCognitive2016; @luiEvidencebasedAppsReview2017; @sucalaAnxietyThereApp2017; @stawarzUserExperienceCognitive2018]. However, few studies on digital self-control have taken this route: To the best of our knowledge, four published studies have described and/or categorised functionality in samples of DSCTs available online [@biskjaerTakingActionDistraction2016; @lyngsSelfControlCyberspaceApplying2019; @mongeroffarelloRaceDigitalWellbeing2019; @vanvelthovenProblematicSmartphoneUse2018]: Velthoven et al. [@vanvelthovenProblematicSmartphoneUse2018] presented aims and features in 21 tools identified via the software recommendation platform [alternativeto.net](https://alternativeto.net); Biskjær et al. [@biskjaerTakingActionDistraction2016] presented a taxonomy of functionality based on 10 tools from the Google Play, Chrome Web, and Apple App stores, and online tech magazines; Monge Roffarello & De Russis [@mongeroffarelloRaceDigitalWellbeing2019] presented a taxonomy of functionality based on 42 mobile apps from the Google Play store; and Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019] presented a taxonomy of functionality based on 367 tools from the Google Play, Chrome Web and Apple App stores, which was mapped to psychological mechanisms in a dual systems model.

Only one of these studies analysed user reviews, namely Monge Roffarello & De Russis [@mongeroffarelloRaceDigitalWellbeing2019], which included thematic analysis of 1,128 reviews from 42 smartphone apps, and none incorporated popularity metrics. Moreover, Monge Roffarello & De Russis' analysis of user reviews was restricted to 42 apps drawn from the Google Play store and did not specify how the reviews were distributed across apps with different functionality. Thematic analysis is likely to yield different results depending on how reviews are sampled across tools with different functionality, and depending on the analysis strategy [@chenARminerMiningInformative2014; @braunThematicAnalysis2018]. Therefore, additional analyses with broader sampling of tools, and transparent sampling of reviews from those tools, might be helpful.

On this background, we extended existing studies of DSCTs by (i) scraping installation numbers, ratings, and user reviews from a broad sample of tools across different stores and types of functionality, (ii) analysing reviews using 'reflexive' thematic analysis [@braunThematicAnalysis2018], (iii) analysing popularity metrics' relationship to design patterns, and (iv) making our materials, analysis scripts and data openly available in the interest of transparency and reusability [@harozUpdatesCHISubmission2019].

@@

Data, analysis scripts, and materials for the present paper are available at <https://osf.io/cdqub/>.

We investigated the 367 digital self-control tools reviewed by Lyngs et al. [@lyngsSelfControlCyberspaceApplying2019], as this paper provided the largest available review of DSCTs. Lyngs et al. identified these tools by downloading search results on the Chrome Web, Google Play, and Apple App stores in August 2018 for the terms 'distraction', 'addiction', 'motivation', and 'self-control', individually as well as prefixed by 'internet', 'smartphone', and 'laptop'. From the initial results (4890 distinct apps and extensions), they included tools explicitly designed to help people self-regulate digital device use, resulting in 367 tools (222 browser extensions from Chrome Web, 87 apps from Google Play, and 58 apps from the Apple App store).[^3]

We adopted Lyngs et al.'s classification of functionality within these tools, which clustered functionality into four types of design patterns for digital self-control:

-   *block/removal* (features for blocking distractions, such as temporarily locking the user out of specific apps, or removing them in the first place, such as by hiding recommended videos on YouTube)

-   *self-tracking* (features for tracking and/or visualising device use, such as showing how much time the user has spent in particular apps or on devices overall, or letting the user start a countdown timer when they need to focus)

-   *goal-advancement* (features for reminding the user of their usage goals, such as displaying pop-ups when a specific amount of time has been spent, or replacing the content of new browser tabs with to-do lists)

-   *reward/punishment* (features that provide incentives for using devices in particular ways, such as rewarding the user with points/streaks when they do not use their phone)

@@

In March 2019, we scraped additional information about these tools, including publicly available user reviews, as well as installation numbers, ratings, and store descriptions. For apps on the Google Play store, we used the open-source `google-play-scraper` [@olanoGoogleplayscraper2018] to scrape reviews and tool information. For apps on the Apple App store, we used the open-source `app-store-scraper` [@olanoAppstorescraper2018] to scrape reviews, and wrote our own R script (using the packages [`rvest`](https://rvest.tidyverse.org) and [`RSelenium`](http://ropensci.github.io/RSelenium/)) to scrape tool information. As the `app-store-scraper` sometimes failed to retrieve all reviews visible on tools' pages on the Apple App store, we wrote an additional R script to scrape reviews when manual inspection found some to be missing. For browser extensions on the Chrome Web store, we wrote our own R script to scrape both tool information and user reviews.

@@

All stores displayed average ratings, as well as a count of how many ratings a tool had received. For installation numbers, the stores provided different information: the Chrome Web store provided an exact number of users, and the Google Play store provided a 'minimum number of installs' (e.g. '100,000+'). The Apple App store provided no direct information about user numbers. For processing that ranked tools by installation numbers, we used number of ratings as a proxy for relative numbers of users for tools from the Apple App store.

## Sampling and thematic analysis of user reviews

To get a broad coverage of reviews for tools across stores and types of design patterns, we wrote an R script to systematically sample user reviews for qualitative analysis from the scraped data. To ensure that the reviews analysed came from tools with a minimum level of engagement on the stores, we only sampled reviews from tools on the Google Play and Chrome Web stores that had at least 100 installations/users, and tools from the Apple App Store that had at least 5 ratings. To make the sampled reviews more likely to be informative, we excluded reviews shorter than 10 words. Our script iterated through the four types of design patterns described above, on each iteration picking the tool on each store that included that specific type of design pattern and which had the highest number of users (Google Play and Chrome Web) or ratings (Apple App store). The script then randomly sampled 30 user reviews from each of those tools (if a tool had fewer than 30 reviews, all of its reviews would be included), and started over. The script ran until at least 1,500 reviews were sampled in total. A specific tool could only have its reviews sampled once.

Next, we conducted inductive thematic analysis of the sampled reviews, following the 'reflexive' approach described in [@braunUsingThematicAnalysis2006] and [@braunThematicAnalysis2018]. Our work was informed by an earlier analysis of 961 reviews randomly sampled from the same data set by the lead author of this paper [@lyngs2021a ch. 4]. However, our analysis in the present paper was conducted from scratch, with a new sample of user reviews drawn in the manner described. After sampling reviews for analysis, the lead author (UL) read through them all and conducted initial coding of recurrent patterns relevant to the research focus. A second author (LC) read through one third of the reviews and independently coded recurrent patterns of meaning. Afterwards, these two authors read through the coded excerpts, recoded them, iteratively sorted the codes into potential themes, and discussed emerging themes.

Thematic coding was conducted using [NVivo](https://www.qsrinternational.com/nvivo-qualitative-data-analysis-software/home/) v1.6.1; all quantitative analyses were conducted in R v4.0.5 (see [osf.io/cdqub/](https://osf.io/cdqub/) for information on R package versions used).

@@

At the time of scraping, `r num_missing` tools from Lyngs et al.'s review had become unavailable. 
Hence, we scraped information for `r nrow(meta_data)` tools: 
`r num_w_info_by_store %>% filter(store == "chrome") %>% pull(n)` extensions from Chrome Web, `r num_w_info_by_store %>% filter(store == "play") %>% pull(n)` apps from Google Play, and `r num_w_info_by_store %>% filter(store == "apple") %>% pull(n)` apps from the Apple App store.

@@

(ref:word-cloud-caption) Word cloud depicting frequencies of terms across all `r format(nrow(reviews_w_main_cat_data), big.mark = ",")` reviews. Font size indicates relative frequency. The word cloud excludes the terms 'app/s', 'phone', and 'extension', as well as stop words (i.e., extremely common words unlikely to be informative such as "the", "of", "to", etc. Stop words were excluded using the R package `tidytext` [@silge2016tidytext]). Colouring is aesthetic and does not map to any characteristics of the data.

@@

```{r, include=FALSE}
(num_reviews_summary <- reviews_w_main_cat_data %>% 
  group_by(tool_id) %>% 
  summarise(num_reviews = n()) %>% 
  summarise(mean = mean(num_reviews),
            min = quantile(num_reviews, 0),
            q25 = quantile(num_reviews, 0.25),
            median = quantile(num_reviews, 0.5),
            q75 = quantile(num_reviews, 0.75),
            max = quantile(num_reviews, 1)))
